---
title: "CUPED: Controlled Experiments Utilizing Pre-Experiment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CUPED: Controlled Experiments Utilizing Pre-Experiment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

CUPED is a method for variance reduction being widely implemented by major 
online experimentation platforms (Deng et al. 2013). The idea was recently 
extended to a general augmentation framework (Deng et al. 2023), which is a 
special case of 

Chen, Y.-H., and Chen, H. (2000), A Unified Approach to Regression Analysis Under Double Sampling Design, *Journal of the Royal Statistical Society, Series B*, 62, 449–460.

The extended CUPED is a good application case of `multipleOutcomes`. Instead of 
following Deng's new paper, I use Chen's approach to elaborate on the details of 
extended CUPED, as this will better facilitate the explanation of its underlying assumptions and failure conditions.

Consider a two-arm randomized controlled experiment of a targeted metric $Y$. Let 
$A$ be the arm assignment. If randomization is successful, the treatment effect 
can be assessed through $\hat{\Delta}$, the estimate of $\Delta$ in the linear 
regression model 

$$Y = \alpha + \Delta A + \text{error},$$

where error is unnecessary to be normal. Note that a logistic regression model 
may be more appropriate if $Y$ is binary.

Obviously, $\hat{\Delta}$ is consistent and asymptotic unbiased under good 
randomization. We are of interest in alternative estimates for $\Delta$ with variance 
lower than that of $\hat{\Delta}$. Consider an estimate 

$$\tilde{\Delta} = \hat{\Delta} + c \times \hat{\delta} $$
where $\text{E}(\hat{\delta}) = 0$. Thus, $\tilde{\Delta}$ is also consistent 
and asymptotic unbiased to the treatment effect $\Delta$, and the variance of 
$\tilde{\Delta}$ is no greater than $\hat{\Delta}$ (worst case: $c=0$). **As long 
as $\hat\Delta$ and $\hat\delta$ are correlated, we can find a $c$ yielding a more 
efficient estimate of $\Delta$.** This simple idea can be further extended for more 
than one $\hat\delta$: 

$$\tilde{\Delta} = \hat{\Delta} + c_1 \times \hat\delta_1 + c_2 \times \hat\delta_2 + \cdots + c_K \times \hat\delta_K.$$

To conduct $\hat\delta$ of zero expectation, one may collect additional pre- or in-experiment metrics $X_1, X_2, \cdots, X_K$ for each of the samples. If $X$ are 
well balanced in the two arms, we have $\delta_1 = \delta_2 = \cdots = \delta_K = 0$, 
where $\delta_k$ is the slope of a working model 

$$X_k = \alpha_k + \delta_k A + \text{error}.$$
Here, a working model means that the relationship between the variables specified 
by the model is unnecessarily true; we simply choose to fit it as is. As a result, 
the error term can be anything and an unreasonable link function may be adopted 
without affecting the validity of extended CUPED. For example, one can fit a linear 
regression against a binary $X$; the slope $\delta = 0$ as long as $X$ is balanced. 
Furthermore, note that for any transformation $h(\cdot)$, $h(X)$ is balanced given 
that $X$ is balanced, one can fit the following working model as well

$$h_k(X_k) = \alpha_k + \delta_k^h A + \text{error}.$$
A more natural model, the logistic regression model also works

$$\text{logit}\ \text{Pr}(X_k = 1\mid A) = \alpha_k + \delta_k^\text{logit}A.$$
It is interesting that mutiple $\hat\delta$ that corresponding to the same $X_k$ 
can be integrated into the extended CUPED estimate simultaneously. In the example 
above, $\tilde\Delta$ can be defined as 

$$\tilde\Delta = \hat\Delta + c_k \hat{\delta}_k + c_k^h \hat{\delta}_k^h + c_k^\text{logit}\hat{\delta}_k^\text{logit} + \hat\delta \text{ based on other } X$$
The variance of this estimate is no greater than an extended CUPED estimate with 
less $\hat\delta$ from the same metric $X_k$ (e.g. $\hat\Delta + c_k \hat\delta_k$).

## Conclusions

In summary, the extended CUPED is valid (consistent and asymptotic 
unbiased) when 

* arm assignment is randomized (good for $\text{E}(\hat\Delta) = \Delta$) **and** 
* pre- and in-experiment metrics are balanced across arms (good for $\text{E}(\hat\delta)=0$).

The extended CUPED is more efficient when 

* at least one of the pre- or in-experiment metrics is correlated with the targeted 
metric **and** 
* the weights $c$ are carefully picked.

The optimal weights $c$ depend on the covariance matrix of $(\hat\Delta, \hat\delta)$, 
not the original observation of $(Y, X)$, thus, tuning on $c$ will not inflate the 
type I error of inference. **`multipleOutcomes` can return the covariance matrix 
for extended CUPED**. 

The extended CUPED remains valid even if 

* the working models for pre- and in-experiment metrics ($X \sim A$) are mis-specified **and/or**
* a pre- or in-experiment metric that is uncorrelated with $Y$ is used in the 
extended CUPED estimate. This is a nice property because we don't have to worry 
about introducing too many irrelevant metrics that are unhelpful. 

In fact, adding multiple $\hat\delta$ based on different transformations on the same $X_k$ into the extended CUPED estimate may lower its variance further in some cases. 
A variance stabilize transformation may be considered for metrics of certain 
parametric families of distributions or highly skewed metrics. 


## Implementation
```{r setup}
## generating data
library(multipleOutcomes)
library(mvtnorm)
library(ggplot2)
library(dplyr)
library(tidyr)

genData <- function(seed = NULL){
  set.seed(seed)
  n <- 500
  sigma <- matrix(.7, 5, 5)
  diag(sigma) <- 1.
  
  x <- rmvnorm(n, sigma = sigma)
  x[, 1] # pre-experiment data of the targeted metric
  x[, 2] # continuous metric
  x[, 3] <- log(abs(x[, 3])) # continuous and skewed metric
  x[, 4] <- ifelse(x[, 4] > .2, 1, 0) # binary metric
  a <- rbinom(n, 1, .5) # 1:1 randomization
  
  # y is post-experiment data of the targeted metric
  y <- x[, 5]
  y[a == 0] <- y[a == 0] + rnorm(sum(a == 0), sd = .1)
  y[a == 1] <- y[a == 1] + rnorm(sum(a == 1), mean = .1, sd = .1) # Delta = .1
  
  ## x5 and x6 are metrics (random noise) but uncorrelated with the targeted metric
  data.frame(
    y, a = a, 
    x1 = x[, 1], x2 = x[, 2], x3 = x[, 3], x4 = x[, 4], 
    x5 = rnorm(n), x6 = rbinom(n, 1, .5))
  
}

dat <- genData(1)
head(dat) %>% print()
xtabs(~x4, dat) %>% print()

plot(
  dat %>% 
    select(-x4, -x6, -a) %>% 
    pivot_longer(everything(), names_to = "indicator", values_to = "value") %>% 
    ggplot(aes(value)) + 
    geom_histogram() + facet_wrap(~indicator)
)

## function to compute CUPED estimate
## ...: formulas for each of the metrics
## family: a vector of family for each of the formulas
## data: a data frame of experiment data
cuped <- function(..., family, data){
  fit <- multipleOutcomes(..., family = family, data = data, data_index = NULL)
  id <- sapply(fit$id_map, function(x){x['a']})
  id1 <- id[1]
  id2 <- id[-1]
  Delta <- coef(fit)[id1]
  delta <- coef(fit)[id2]
  mcov <- vcov(fit)
  opt_c <- as.vector(- solve(mcov[id2, id2, drop = FALSE]) %*% mcov[id2, id1, drop = FALSE])
  
  estimate <- Delta + sum(delta * opt_c)
  stderr <- (
    mcov[id1, id1] + 
    2 * as.vector(t(opt_c) %*% mcov[id2, id1, drop = FALSE]) + 
    as.vector(t(opt_c) %*% mcov[id2, id2, drop = FALSE] %*% opt_c)
  ) %>% sqrt()
  
  list(estimate = estimate, stderr = stderr)
}
```

## An Example
```{r}
## An example
## y ~ a: treatment effect of post-experiment data for the targeted metric
## x1 ~ a: delta (= 0) using pre-experiment data for the targeted metric
## x2 ~ a: delta (= 0) using a continuous pre-experiment metric
## x3 ~ a: delta (= 0) using a skewed pre-experiment metric
## abs(x3) ~ a: delta (= 0) using a transformation of a pre-experiment metric
## x4 ~ a: delta (= 0) using a binary pre-experiment metric
## x5 ~ a: delta (= 0) using a metric that is a random noise
## family: linear regression model is fitted for all working models (even if x4 is binary)
cuped_fit <- 
  cuped(
    y ~ a, 
    x1 ~ a, 
    x2 ~ a, 
    x3 ~ a, 
    abs(x3) ~ a, 
    x4 ~ a, 
    x5 ~ a, 
    family = 'gaussian', 
    data = dat)

## CUPED estimate and its standard error
print(cuped_fit)

## treatment effect on targeted metric (post-experiment data only)
summary(lm(y ~ a, dat))$coef

```
The example above illustrates a complex case of CUPED using `mutipleOutcomes`. 
The standard error of estimated treatment effect is substantially reduced 
(`r cuped_fit['stderr'] %>% sprintf('%.3f', .)` vs `r summary(lm(y ~ a, dat))$coef['a', 'Std. Error'] %>% sprintf('%.3f', .)`).

## Simulation

We carry out simulations to verify the key points in Conclusions. Specifically, 

* adding more metrics with zero expectation can reduce variance;
* adding noise does not hurt the efficiency; 
* if all added metrics are noise, CUPED is equivalent to using post-data only.

```{r}
n_simu <- 1000
all_estimate <- NULL
all_stderr <- NULL
for(i in 1:n_simu){
  dat <- genData(i)
  
  # use post- data only
  post_only <- summary(lm(y ~ a, dat))$coef
  
  # add pre- data of targeted metric
  pre_target <- cuped(y ~ a, x1 ~ a, family = 'gaussian', data = dat)
  
  # add more pre- data
  pre_more <- cuped(y ~ a, x1 ~ a, x3 ~ a, x4 ~ a, family = 'gaussian', data = dat)
  
  # multiple delta (transformation) for the same metric
  # transform_delta <- cuped(y ~ a, x1 ~ a, x3 ~ a, exp(x3) ~ a, x4 ~ a, x4 ~ a, family = c('gaussian', 'gaussian', 'gaussian', 'gaussian', 'gaussian', 'binomial'), data = dat)
  
  # add all correlated pre- data
  all_corr <- cuped(y ~ a, x1 ~ a, x2 ~ a, x3 ~ a, x4 ~ a, family = 'gaussian', data = dat)
  
  # add noise data
  add_noise <- cuped(y ~ a, x1 ~ a, x2 ~ a, x3 ~ a, x4 ~ a, x5 ~ a, x6 ~ a, family = 'gaussian', data = dat)
  
  # noise only
  noise_only <- cuped(y ~ a, x5 ~ a, x6 ~ a, family = c('gaussian', 'gaussian', 'binomial'), data = dat)
  
  extractEstimate <- function(mod){
    mod$estimate
  }
  
  extractStderr <- function(mod){
    mod$stderr
  }
  
  estimate <- 
    data.frame(
      post_only = post_only['a', 'Estimate'], 
      pre_target = extractEstimate(pre_target), 
      pre_more = extractEstimate(pre_more), 
      # transform_delta = extractEstimate(transform_delta), 
      all_corr = extractEstimate(all_corr), 
      add_noise = extractEstimate(add_noise), 
      noise_only = extractEstimate(noise_only)
    )
  
  stderr <- 
    data.frame(
      post_only = post_only['a', 'Std. Error'], 
      pre_target = extractStderr(pre_target), 
      pre_more = extractStderr(pre_more), 
      # transform_delta = extractStderr(transform_delta), 
      all_corr = extractStderr(all_corr), 
      add_noise = extractStderr(add_noise), 
      noise_only = extractStderr(noise_only)
    )
  
  all_estimate <- rbind(all_estimate, estimate)
  all_stderr <- rbind(all_stderr, stderr)

}

## mean of estimates of simulation (true Delta = .1)
apply(all_estimate, 2, mean) %>% print(digits = 3)

## empirical standard deviation
apply(all_estimate, 2, sd) %>% print(digits = 3)

## theoretical standard error
apply(all_stderr, 2, mean) %>% print(digits = 3)
```
